Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population

### 2025-11-25 manuscript/manuscript.qmd
adding manuscript revisions

---
title: "Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population"
date: "2025-09-25"
author:
  - name: Cormac Monaghan
    url: https://c-monaghan.github.io/
    email: cormacmonaghan@protonmail.com
    orcid: 0000-0001-9012-3060
    affiliation:
      - id: 1
        name: Maynooth University
    attributes:
      corresponding: true
  - name: Rafael de Andrade Moral
    url: https://rafamoral.github.io/
    orcid: 0000-0002-0875-3563
    affiliation:
      - id: 1
        name: Maynooth University
  - name: Michelle Kelly
    url: https://www.ncirl.ie/About/A-Z-Staff-Directory/Staff/346
    orcid: 0000-0001-6159-6707
    affiliation:
      id: 2
      name: National College of Ireland
  - name: Joanna McHugh Power
    url: https://www.maynoothuniversity.ie/people/joanna-mchugh-power
    orcid: 0000-0002-7387-3107
    affiliation:
      id: 1
      name: Maynooth University
correspondence-prefix: "Corresponding author"
abstract: |
  Cognitive decline is a global health concern, making the identification of early, modifiable risk factors essential. While apathy is a recognized prodromal marker, procrastination may also signal early executive dysfunction. We used longitudinal secondary data from the United States Health and Retirement Study among adults aged 60+ $(n = 549; \bar{x} = 69.70;\sigma = 7.58)$. Cognitive function, procrastination, and a proxy measure of apathy were assessed. Transitions between normative cognitive function, mild cognitive impairment (MCI), and dementia were modeled using a discrete-time first-order Markov model. Procrastination scores were higher among individuals with MCI or dementia than those with normative cognitive function. Procrastination also interacted with age, disproportionately increasing the risk of decline in the oldest participants. Procrastination was associated with cognitive impairment and predicted transitions to MCI, suggesting it may serve as both an early behavioral marker and compounding risk factor.
additional-info: |
  **Data Availability:** All data and replication code are publicly available at <https://github.com/C-Monaghan/crisp-clam>
keywords:
  - Dementia risk
  - Mild cognitive impairment
  - Procrastination
  - Apathy
  - Executive dysfunction
  - Aging
published: "This manuscript is a preprint and is currently under review at *Alzheimer's & Dementia: Diagnosis, Assessment & Disease Monitoring*"
code-repo: "Access the code, data, and analysis at <https://github.com/C-Monaghan/crisp-clam>"
bibliography: references.bib
csl: csl/apa.csl
block-author: false
link-citations: true
editor:
  markdown:
    wrap: 80
keep-tex: true
---

@@

The Lancet Commission on Dementia identified fourteen causal risk factors
including smoking, hypertension, and diabetes, which account for approximately
45% of dementia cases worldwide [@livingston2024]. Prodromal markers, on the
other hand, reflect early disease processes and may signal limited opportunities
for prevention [@teipel2025]. One such prodromal marker is apathy, or the
significant loss of motivation, as distinct from depression and cognitive
impairment [@fresnais2023; @richard2012]. Apathy is prevalent in both MCI and
dementia subgroups [@donovan2014; @fresnais2023, @salem2023; @vandalen2018].
Apathy is a significant predictor of the transition from MCI to dementia
[@ruthirakuhan2019; @salem2023, @vandalen2018]. While apathy has been
well-established as a prodromal marker and risk factor for dementia progression,
emerging evidence suggests that procrastination may also relate to early
cognitive decline.

Though superficially similar, apathy and procrastination are behaviorally
distinct. Apathy reflects reduced internal drive and emotional engagement,
impairing initiation of action [@fresnais2023], whereas procrastination reflects
intact intention but delayed execution [@steel2007]. While both reflect
impairments in executive functioning, procrastination may additionally capture
unique aspects of emotional and motivational dysregulation . For instance,
person A may feel no motivation to attend an exercise class and therefore never
books it (apathy). Person B, however, books the class and intends to go, but
postpones at the last minute (procrastination). Both apathy and procrastination
have been linked to dysfunction in prefrontal regions implicated in dementia
[@friden2020; @zhang2010; @joseph2021; @fahed2021].

@@

We focused on four waves of HRS data from 2016 to 2022. Specifically, our study
sample consisted of respondents who participated in an experimental module
assessing procrastination during the 2020 wave. These experimental modules,
administered at the end of the core HRS interview, consist of concise
questionnaires designed to explore new topics or supplement existing core survey
data [@juster1995]. Each respondent receives only one experimental module, with
sample sizes for each module constituting approximately 10% of the core sample.
As a result, while the core HRS sample includes approximately $18,000$
respondents, our initial sample of interest consisted of $1,368$ respondents. We
excluded respondents with missing cognitive assessment data for any wave
$(n = 419)$, those under 60 years of age, (as cognitive symptoms typically occur
around this age $n = 398$), and those with complete missing values across the
procrastination measure $(n = 2)$. This resulted in a final analytic sample of
$549$ respondents.

@@

While no direct measure of apathy exists within the HRS, we utilised two
questions from the eight-item version of the Center for Epidemiological Studies
Depression (CES-D8) scale [@briggs2018] as proxies for apathy: "You felt that
everything you did was an effort" and "You could not get going". Both items
capture core features of apathy (behavioural and motivational disengagement)
and, while not a comprehensive measure of apathy, provide a valid and pragmatic
approximation for modelling purposes. Each item was measured on a binary
“yes/no” scale with total scores ranging from \$0\$ to \$2\$.

@@

\clearpage

@@

# Table information
tbl_descriptives <- tibble::tribble(
  ~""                 , ~Full              , ~NC                , ~MCI              , ~Dementia        ,
  "Age (years)"       , "69.70 ± 7.58"    , "69.70 ± 7.55"    , "69.60 ± 7.55"   , "70.00 ± 9.89"  ,
  "Male"              , "38.43% (n = 211)" , "39.82% (n =180)"  , "32.56% (n = 28)" , "27.27% (n = 3)" ,
  "Female"            , "61.57% (n = 338)" , "60.18% (n = 272)" , "67.44% (n = 58)" , "72.73% (n = 8)" ,
  "No degree"         , "16.24% (n = 98)"  , "10.56% (n = 47)"  , "39.53% (n = 34)" , "63.54% (n = 7)" ,
  "GED"               , "51.58% (n = 279)" , "51.91% (n = 231)" , "52.33% (n = 45)" , "27.27% (n = 3)" ,
  "Further education" , "32.29% (n = 175)" , "37.53% (n = 167)" , "8.14% (n = 7)"   , "9.09% (n = 1)"  ,
  "Apathy"            , "0.37 ± 0.63"     , "0.31 ± 0.59"     , "0.65 ± 0.76"    , "0.36 ± 0.51"   ,
  "Procrastination"   , "28.60 ± 12.00"   , "27.70 ± 11.30"   , "32.10 ± 13.80"  , "39.00 ± 16.00"
)

@@

# Outputting table
tbl_descriptives |>
  setNames(c(" ", "Full sample (n = 549)", "NC (n = 452)", "MCI (n = 86)", "Dementia (n = 11)")) |>
  tt(notes = list(
    caption,
    a = list(i = 10, j = 1, text = footnote))
  ) |>
   group_tt(i = list("Sex" = 2, "Education" = 4)) |>
   style_tt(j = 2:5, align = "c") |>
   style_tt(i = c(2, 5), align = "l", bold = TRUE) |>
   format_tt(escape = TRUE) |>
   format_tt("notes", markdown = TRUE) |>
   style_tt(
     bootstrap_class = "table table-sm"
     )
```

@@

Despite the insights offered by this study, several limitations should be
considered. Most notably, although our longitudinal Markov modelling approach
captured changes in cognitive status, a key constraint was the measurement of
procrastination at only a single time point (2020). As a result, we assumed
temporal stability in procrastination scores across time and used this measure
as both a retrospective and prospective proxy for procrastination scores.
However, this precluded analysis of within-person changes in this behavior over
time and whether such changes affect cognitive transitions. It should be noted
that this limitation is inherent to the HRS dataset rather than a methodological
oversight (Juster & Suzman, 1995). Future studies should prioritise repeated
assessments of procrastination to determine whether increases in procrastination
precedes, accompanies, or follows cognitive decline. Moreover, our use of a
proxy measure of apathy may not have fully captured the multidimensional nature
of apathy. Future research should incorporate more validated apathy assessments
to strengthen the behavioral inferences drawn.

Additionally, as noted earlier, the relatively small number of individuals in
the MCI and dementia groups may have constrained statistical power for certain
comparisons and increased the potential for misclassification bias. Finally,
while the models adjusted for several key demographic and psychological
covariates, unmeasured confounding (undiagnosed medical conditions, medication
use, or sleep quality) could also have influenced cognitive outcomes. Future
research should seek to replicate and extend these findings using more
clinically diverse samples.

>>>>

---
title: "Procrastination as a Marker of Cognitive Decline: Evidence from Longitudinal Transitions in the Older Adult Population"
date: "2025-11-25"
author:
  - name: Cormac Monaghan
    url: https://c-monaghan.github.io/
    email: cormacmonaghan@protonmail.com
    orcid: 0000-0001-9012-3060
    affiliation:
      - id: 1
        name: Maynooth University
    attributes:
      corresponding: true
  - name: Rafael de Andrade Moral
    url: https://rafamoral.github.io/
    orcid: 0000-0002-0875-3563
    affiliation:
      - id: 1
        name: Maynooth University
  - name: Michelle Kelly
    url: https://www.ncirl.ie/About/A-Z-Staff-Directory/Staff/346
    orcid: 0000-0001-6159-6707
    affiliation:
      id: 2
      name: National College of Ireland
  - name: Joanna McHugh Power
    url: https://www.maynoothuniversity.ie/people/joanna-mchugh-power
    orcid: 0000-0002-7387-3107
    affiliation:
      id: 1
      name: Maynooth University
correspondence-prefix: "Corresponding author"
abstract: |
  Cognitive decline is a global health concern, making the identification of early, modifiable risk factors essential. While apathy is a recognized prodromal marker, procrastination may also signal early executive dysfunction. We used longitudinal secondary data from the United States Health and Retirement Study among adults aged 60+ $(n = 549; \bar{x} = 69.70;\sigma = 7.58)$. Cognitive function, procrastination, and a proxy measure of apathy were assessed. Transitions between normative cognitive function, mild cognitive impairment (MCI), and dementia were modeled using a discrete-time first-order Markov model. Procrastination scores were higher among individuals with MCI or dementia than those with normative cognitive function. Procrastination also interacted with age, disproportionately increasing the risk of decline in the oldest participants. Procrastination was associated with cognitive impairment and predicted transitions to MCI, suggesting it may serve as both an early behavioral marker and compounding risk factor.
additional-info: |
  **Data Availability:** All data and replication code are publicly available at <https://github.com/C-Monaghan/crisp-clam>
keywords:
  - Dementia risk
  - Mild cognitive impairment
  - Procrastination
  - Apathy
  - Executive dysfunction
  - Aging
published: "This manuscript is a preprint and is currently under review at *Alzheimer's & Dementia: Diagnosis, Assessment & Disease Monitoring*"
code-repo: "Access the code, data, and analysis at <https://github.com/C-Monaghan/crisp-clam>"
bibliography: references.bib
csl: csl/apa.csl
block-author: false
link-citations: true
editor:
  markdown:
    wrap: 80
keep-tex: true
---

@@

The Lancet Commission on Dementia identified fourteen causal risk factors
including smoking, hypertension, and diabetes, which account for approximately
45% of dementia cases worldwide [@livingston2024]. Prodromal markers, on the
other hand, reflect early disease processes and may signal limited opportunities
for prevention [@teipel2025]. One such prodromal marker is apathy, or the
significant loss of motivation, as distinct from depression and cognitive
impairment [@fresnais2023; @richard2012]. Apathy is prevalent in both MCI and
dementia subgroups [@donovan2014; @fresnais2023; @salem2023; @vandalen2018].
Apathy is a significant predictor of the transition from MCI to dementia
[@ruthirakuhan2019; @salem2023; @vandalen2018]. While apathy has been
well-established as a prodromal marker and risk factor for dementia progression,
emerging evidence suggests that procrastination may also relate to early
cognitive decline.

Although apathy and procrastination can appear similar in that both lead to
reduced engagement in everyday activities, the underlying behavioural processes
are distinct. Apathy reflects reduced internal drive and emotional engagement,
impairing initiation of action [@fresnais2023]. In contrast, procrastination is
a self-regulatory failure defined by the voluntary delay of an important and
intended course of action despite expecting negative consequences [@sirois2013;
@klingsieck2013]. This distinction suggests that individuals who procrastinate
typically form intentions but fail to execute them in a timely manner. This
failure can often be understood through an affective framework, where delayed
action may stem from the avoidance of negative emotions (i.e., stress or
anxiety) associated with the task [@sirois2013]. Furthermore, while also
conceptualised as a stable trait linked to specific personality factors
[@meng2024; @sirois2017c], these factors can shift in later life and may
interact with emerging cognitive or motivational changes to influence
procrastination behaviour in older adult [@oi2024; @donnellan2008].

Both apathy and procrastination have been linked to dysfunction in prefrontal
regions implicated in dementia [@friden2020; @zhang2010; @joseph2021;
@fahed2021]. Procrastination has been associated with alterations in brain
circuitry supporting cognitive control and emotional regulation. Research has
identified both structural and functional differences primarily in the
prefrontal cortex (PFC) and parahippocampal cortex [@zhang2019]. Specifically,
high procrastination is associated with decreased gray matter volume in the
dorsolateral PFC (dlPFC) and increased spontaneous activity in the ventromedial
PFC (vmPFC) [@liu2017]. Additionally, functional connectivity studies indicate
altered couplings between both the vmPFC and dlPFC, along with the vmPFC and
frontal gyri [@zhang2019]. Critically, these same networks are affected in
neurodegenerative conditions, including Alzheimer’s disease [@joseph2021].

@@

We focused on four waves of HRS data from 2016 to 2022. Specifically, our study
sample consisted of respondents who participated in an experimental module
assessing procrastination during the 2020 wave. These experimental modules,
administered at the end of the core HRS interview, consist of concise
questionnaires designed to explore new topics or supplement existing core survey
data [@juster1995]. Each respondent receives only one experimental module, with
sample sizes for each module constituting approximately 10% of the core sample.
As a result, while the core HRS sample includes approximately $18,000$
respondents, our inclusion criteria (i.e., having completed the experimental
module) identified an initial sample of $1,368$ respondents. We then applied the
following exclusion criteria: respondents with missing cognitive assessment data
for any $(n = 419)$, those under 60 years of age, (as cognitive symptoms
typically occur around this age $n = 398$), and those with complete missing
values across the procrastination measure $(n = 2)$. The application of these
exclusion criteria resulted in a final analytic sample of $549$ respondents, of
whom $61.57\% (n = 338)$ were female.

@@

While no direct measure of apathy exists within the HRS, we utilised two
questions from the eight-item version of the Center for Epidemiological Studies
Depression (CES-D8) scale [@briggs2018] as proxies for apathy: "You felt that
everything you did was an effort" and "You could not get going". Both items
capture core features of apathy (behavioural and motivational disengagement)
and, while not a comprehensive measure of apathy, provide a valid and pragmatic
approximation for modelling purposes. Each item was measured on a binary
“yes/no” scale with total scores ranging from $0$ to $2$.

@@

@@

# Table information
tbl_descriptives <- tibble::tribble(
  ~""                 , ~Full              , ~NC                , ~MCI              , ~Dementia        ,
  "Age (years)"       , "69.70 ± 7.58"    , "69.70 ± 7.55"    , "69.60 ± 7.55"   , "70.00 ± 9.89"  ,
  "Male"              , "38.43%<br>(n = 211)" , "39.82%<br>(n = 180)"  , "32.56%<br>(n = 28)" , "27.27%<br>(n = 3)" ,
  "Female"            , "61.57%<br>(n = 338)" , "60.18%<br>(n = 272)" , "67.44%<br>(n = 58)" , "72.73%<br>(n = 8)" ,
  "No degree"         , "16.24%<br>(n = 98)"  , "10.56%<br>(n = 47)"  , "39.53%<br>(n = 34)" , "63.54%<br>(n = 7)" ,
  "GED"               , "51.58%<br>(n = 279)" , "51.91%<br>(n = 231)" , "52.33%<br>(n = 45)" , "27.27%<br>(n = 3)" ,
  "Further education" , "32.29%<br>(n = 175)" , "37.53%<br>(n = 167)" , "8.14%<br>(n = 7)"   , "9.09%<br>(n = 1)"  ,
  "Apathy"            , "0.37 ± 0.63"     , "0.31 ± 0.59"     , "0.65 ± 0.76"    , "0.36 ± 0.51"   ,
  "Procrastination"   , "28.60 ± 12.00"   , "27.70 ± 11.30"   , "32.10 ± 13.80"  , "39.00 ± 16.00"
)

@@

# Outputting table
tbl_descriptives |>
  setNames(c(" ", "Full sample<br>(n = 549)", "NC<br>(n = 452)", "MCI<br>(n = 86)", "Dementia<br>(n = 11)")) |>
  tt(notes = list(caption,
  a = list(i = 10, j = 1, text = footnote))
  ) |>
   group_tt(i = list("Sex" = 2, "Education" = 4)) |>
   style_tt(j = 2:5, align = "c") |>
   style_tt(j = 1, align = "l") |>
   style_tt(i = c(2, 5), align = "l", bold = TRUE) |>
   # format_tt(escape = TRUE) |>
   format_tt(linebreak = "<br>") |>
   format_tt("notes", markdown = TRUE) |>
   style_tt(
     bootstrap_class = "table table-sm"
     )
```

@@

Despite the insights offered by this study, several limitations should be
considered. Most notably, although our longitudinal Markov modelling approach
captured changes in cognitive status, a key constraint was the measurement of
procrastination at only a single time point (2020). As a result, we assumed
temporal stability in procrastination scores across time and used this measure
as both a retrospective and prospective proxy for procrastination scores.
However, this precluded analysis of within-person changes in this behavior over
time and whether such changes affect cognitive transitions. It should be noted
that this limitation is inherent to the HRS dataset rather than a methodological
oversight (Juster & Suzman, 1995). Future studies should prioritise repeated
assessments of procrastination to determine whether increases in procrastination
precedes, accompanies, or follows cognitive decline.

A further limitation concerns the classification of cognitive status. As our
analyses rely on the HRS, we are constrained by its operational definitions of
MCI and dementia [@crimmins2011], which are based on threshold scores on the
TICS. Although widely used [@langa2017; @davydow2015; @wei2018], these cut-offs
may not fully align with other clinical diagnostic criteria or
neuropsychological assessments, raising the possibility of misclassification.
Future research using more comprehensive cognitive batteries or clinical
diagnoses would help validate and refine the patterns observed here.

Additionally, negative emotional states (e.g., anxiety or depression) are
important psychological factors linked to both cognitive ageing and
procrastination. However, we were unable to adjust for anxiety in our models due
to substantial missingness in the HRS anxiety measure (approximately 61% missing
for our sample). Similarly, depression was not included as a covariate because
two CES-D items were used to construct our proxy measure of apathy, leaving a
non-standard six-item score that would not provide a valid index of depressive
symptoms. Future research using datasets with more complete emotional-health
assessments may help clarify the role of negative emotions in these
associations.

sMoreover, our use of a proxy measure of apathy may not have fully captured the
multidimensional nature of apathy. Future research should incorporate more
validated apathy assessments to strengthen the behavioral inferences drawn.
Additionally, as noted earlier, the relatively small number of individuals in
the MCI and dementia groups may have constrained statistical power for certain
comparisons and increased the potential for misclassification bias. Finally,
while the models adjusted for several key demographic and psychological
covariates, unmeasured confounding (undiagnosed medical conditions, medication
use, or sleep quality) could also have influenced cognitive outcomes. Future
research should seek to replicate and extend these findings using more
clinically diverse samples.
